\chapter{Finance Theory (Title)}
\label{chap:finance_theory}

Yield curve modeling sits at the intersection of stochastic analysis and financial theory. This chapter develops both, following \cite{Bjork2020} for the mathematical foundations of stochastic calculus and the no-arbitrage pricing framework, and \textcolor{red}{reference} for the theory of interest rates, culminating in the pricing of zero coupon bonds and the construction of the yield curve.


\section{Stochastic Calculus (title?)}

Consider a continuous time setting with time $t\in[0,T]$, where the processes of interest are modeled as diffusion processes, denoted $X$ throughout. Informally, a stochastic process is a diffusion if its evolution over a small time interval $[t,t+\Delta t]$ can be approximated by 
$$X_{t+\Delta t}-X_t=\mu(t,X_t)\Delta t +\sigma (t,X_t)Z_t$$
where $\mu(t,X_t)$ a locally deterministic drift term, $Z_t$ is a Gaussian disturbance independent of everything that has happened up to time $t$, and $\sigma(t,X_t)$ is the diffusion term scaling the magnitude of the disturbance. The Gaussian disturbance $Z_t$ is modeled through the Brownian motion, defined below.
\begin{definition}
A stochastic process $W$ is a Brownian Motion if the following conditions hold
\begin{enumerate}
    \item $W_0=0$
    \item The process $W$ has independent increments, i.e. if $r<s\leq t <u$ then $W_u-W_t$ and $W_s-W_r$ are independent random variables.
    \item For $s<t$ the random variable $W_t-W_s$ is Gaussian distributed with  mean $0$ and variance $t-s$, i.e. $W_t-W_s\sim\mathcal{N}(0,t-s)$.
    \item $W$ has continuous trajectories.
\end{enumerate}
\label{def:BM}
\end{definition}


Since the stochastic process evolves over time, we must keep track of the information available at each point in time. The information generated by observing $W$ up to time $t$ is captured by the natural filtration $\mathcal{F}_t^W=\sigma(W_s;0\leq s\leq t)$. A stochastic process $X$ is said to be adapted to the filtration $\{\mathcal{F}_t^W\}_{t\geq 0}$ if $X_t$ is $\mathcal{F}_t^W$-measurable for all $t\geq 0$, meaning the value of $X_t$ depends only on the history of $W$ up to time $t$.


The diffusion processes mentioned above are formally defined through stochastic differential equations (SDEs), which describe the infinitesimal evolution of $X$ over time. Taking the continuous time limit, the discrete approximation above gives rise to an SDE which in integral form reads
\begin{equation}
    X_t=x_0+\int_0^t\mu_s(X_s)ds+\int_0^t\sigma_s(X_s)dW_s
    \label{eq:SDEIntegral}
\end{equation}
 or equivalently in differential form
\begin{align}
    dX_t &= \mu_t(X_t)dt+\sigma_t(X_t)dW_t \label{eq:SDE}\\
    X_0&=x_0 \nonumber
\end{align}
where $x_0$ is the initial value of the process. Both representations involve a stochastic integral with respect to $W$, the precise meaning of which we establish next.



\subsection{The Stochastic Integral}
We introduce the stochastic integral, also known as the Itô integral,
\begin{equation}
    \int_a^bg_sdW_s
    \label{ItoIntegral}
\end{equation}
which is well-defined for $g$ belonging to the class $\mathcal{L}^2$, defined as follows.

\begin{definition}
    Let $\mathcal{L}^2=\mathcal{L}^2([a,b])$ denote the class of stochastic processes $g$ satisfying:
    \begin{enumerate}
        \item $\int_a^b E[g_s^2]\, ds < \infty$.
        \item $g$ is adapted to the $\mathcal{F}_t^W$-filtration.
    \end{enumerate}
    \label{def:L2}
\end{definition}

The stochastic integral $\int_a^bg_sdW_s$ is well-defined for all $g\in\mathcal{L}^2$, and satisfies the following properties.

\begin{proposition}
    Let $g\in\mathcal{L}^2$. Then
    \begin{enumerate}[label=(\alph*)]
        \item Zero mean:
        \begin{equation}
            E\left[\int_a^b g_s\, dW_s \right] = 0
        \end{equation}
        \item Itô isometry:
        \begin{equation}
            E\left[\left(\int_a^b g_s\, dW_s\right)^2\right] = \int_a^b E[g_s^2]\, ds
        \end{equation}
        \item Measurability: The stochastic integral $\int_a^b g_s\, dW_s$ is $\mathcal{F}_b^W$-measurable.
        \item Conditional zero mean: For any $g\in\mathcal{L}^2[s,t]$
        \begin{equation}
            E\left[\int_s^t g_u\, dW_u \,\middle|\, \mathcal{F}_s^W\right] = 0.
        \end{equation}
    \end{enumerate}
    \label{prop:SIproperties}
\end{proposition}







\subsection{Martingales}

An immediate consequence of Proposition \ref{prop:SIproperties} is that stochastic integrals are martingales.

\begin{definition}[Martingale]
    A stochastic process $X$ is called an $(\mathcal{F}_t)$-martingale if the following conditions hold:
    \begin{enumerate}
        \item $X$ is adapted to the filtration $\{\mathcal{F}\}_{t\geq0}$.
        \item For all $t$ we have 
        $$E[|X_t|]<\infty$$
        \item For all $s$ and $t$ with $s\leq t$ the following relation holds
        $$E[X_t|\mathcal{F}_s]=X_s.$$
    \end{enumerate}
    \label{def:martingale}
\end{definition}

The defining feautre of a martingale is the absence of drift. Given the information available at time $s$, the expected value of $X_t$ is simply its current value $X_s$. The following corollary establishes that stochastic integrals are martingales. 

\begin{corollary}
    For any $g\in\mathcal{L}^2$, the process $X_t=\int_0^tg_sdW_s$ is a $\{\mathcal{F}_t^W\}$-martingale.
    \label{cor:SImartingale}
\end{corollary}

\begin{proof}
Let $g\in\mathcal{L}^2$. Since $X_t=\int_0^tg_sdW_s$ is well-defined, we split the integral at $s\leq t$ as
\begin{align*}
    X_t=\int_0^sg_udW_u+\int_s^tg_udW_u=X_s+\int_s^tg_udW_u
\end{align*}
Taking conditional expectations with respect to $\mathcal{F}_s^W$ on both sides applying property $(d)$ of Proposition \ref{prop:SIproperties} gives
$$E[X_t\,|\, \mathcal{F}_s^W]=X_s+E\left[\int_s^t g_u\, dW_u\,\middle|\,\mathcal{F}_s^W\right]=X_s$$
which satisfies condition (3) of Definition \ref{def:martingale}. Adaptiveness and integrability follow directly from $g\in\mathcal{L}^2$, and hence $X_t$ is a $\{\mathcal{F}_t^W\}$-martingale.
\end{proof}

Any stochastic integral $\int_0^tg_sdW_s$ with $g\in\mathcal{L}^2$ is therefore a martingale. Recalling the SDE $\eqref{eq:SDE}$, this means that when the drift coefficient $\mu(t,X_t)$ vanishes, the process $X_t$ itself becomes a martingale.



\subsection{Itô's Formula}

Itô's formula is a central result of stochastic calculus, providing the stochastic differential of any smooth transformation $f(t,X_t)$ of a diffusion process. To state it in full generality, we consider a $n$-dimensional process $X=(X^1,...,X^n)^\top$ driven by $d$ Brownian motions $W=(W^1,...,W^d)^\top$ with instantaneous correlation structure
\begin{equation}
    dW^i_t \cdot dW^j_t = \rho_{ij}\,dt,
\end{equation}
where $\rho$ is the correlation matrix.

The dynamics of $X$ are given by
\begin{equation}
    d\underbrace{\begin{bmatrix}
    X_t^1\\ \vdots\\ X_t^n
    \end{bmatrix}}_{n\times 1}=
    \underbrace{\begin{bmatrix}
    \mu_t^1\\ \vdots\\ \mu_t^n
    \end{bmatrix}}_{n\times 1}dt+
    \underbrace{\begin{bmatrix}
    \sigma_t^{11} & \sigma_t^{12} & \dots & \sigma_t^{1d}\\
    \sigma_t^{21} & \sigma_t^{22} & \dots & \sigma_t^{2d}\\
    \vdots & \vdots & \ddots & \vdots\\
    \sigma_t^{n1} & \sigma_t^{n2} & \dots & \sigma_t^{nd}\\
    \end{bmatrix}}_{n\times d}
    d\underbrace{\begin{bmatrix}
    W^1_t\\ \vdots\\ W^d_t
    \end{bmatrix}}_{d\times 1}
\label{eq:multiSDE}
\end{equation}
which is a multidimensional generalization of \eqref{eq:SDE}, written compactly as $dX_t=\mu_tdt+\sigma_tdW_t$, where each component satisfies
$$dX_t^i=\mu^i_tdt+\sum_{j=1}^d\sigma^{ij}_tdW_t^j.$$
For notational convenience, the dependence of the drift $\mu_t\in\mathbb{R}^n$ and the diffusion matrix $\sigma_t\in\mathbb{R}^{n\times d}$ on the current state $X_t$ is suppressed throughout, with the understanding that $\mu_t=\mu_t(X_t)$ and $\sigma_t=\sigma_t(X_t)$ as in \eqref{eq:SDE}.



\begin{theorem}[Itô's Formula]
Let the $n$-dimensional process $X$ have dynamics given by \eqref{eq:multiSDE} and let $f \in C^{1,2}(\mathbb{R}_+ \times \mathbb{R}^n)$. Then $f(t,X_t)$ \textcolor{red}{(is itself an Itô process with)} has a stochastic differential given by
\begin{equation}
    df(t,X_t) = \left\{
        \frac{\partial f}{\partial t}
        + \sum_{i=1}^n \mu_t^i \frac{\partial f}{\partial x^i}
        + \frac{1}{2}\,\mathrm{tr}\!\left[\sigma_t^\top H_f \sigma_t\right]
    \right\}dt
    + \sum_{i=1}^n \frac{\partial f}{\partial x^i}\,\sigma_t^i\,dW_t,
    \label{eq:Ito}
\end{equation}
where $tr$ denotes the trace of a matrix, and $H$ denotes the Hessian matrix with
$$H_{ij}=\frac{\partial^2f}{\partial {x^i}\partial {x^j}}.$$
\label{thm:ito}
\end{theorem}

\begin{remark}[Itô Multiplication Table]
Theorem~\ref{thm:ito} follows from the formal multiplication rules
\begin{equation}
\begin{cases}
    (dt)^2 = 0,\\
    dt \cdot dW^i_t = 0,\\
    dW^i_t \cdot dW^j_t = \rho_{ij}\,dt,
\end{cases}
\end{equation}
where the last rule reduces to $dW^i_t \cdot dW^j_t = 0$ for $i \neq j$ 
in the independent case.
\label{rem:multitable}
\end{remark}

\textcolor{red}{
\begin{remark}[Infinitesimal Generator \textcolor{red}{(1dim)}]
The drift in \eqref{eq:Ito} decomposes naturally into an explicit time 
derivative $\partial_t f$ and a purely spatial part determined by 
$\mu$ and $\sigma$. Defining the latter as the infinitesimal generator
\begin{equation}
    \mathcal{A}f(x) = \sum_{i=1}^n \mu_i\frac{\partial f}{\partial x_i}
    + \frac{1}{2}\sum_{i,j=1}^n (\sigma\sigma^\top)_{ij}
    \frac{\partial^2 f}{\partial x_i \partial x_j},
\end{equation}
Itô's formula takes the compact form 
$df = (\partial_t f + \mathcal{A}f)\,dt + \nabla_x f \cdot \sigma\,dW_t$.
In particular, since the stochastic integral has zero mean by 
Proposition~\ref{prop:SIproperties}, $f(t,X_t)$ is a martingale if and 
only if $\partial_t f + \mathcal{A}f = 0$. This connection between the dynamics of $X$ and the theory of partial differential equations is made precise by the Feynman-Kac formula of Section~\ref{subsec:fk}, and underlies the pricing framework developed in later chapters.
\end{remark}
}

\textcolor{red}{
\begin{remark}[Infinitesimal Generator \textcolor{red}{(multidim)}]
The drift in \eqref{eq:Ito} decomposes naturally into an explicit 
time derivative $\partial_t f$ and a purely spatial part determined by 
$\mu_t$ and $\sigma_t$. Defining the latter as the infinitesimal generator
\begin{equation}
    \mathcal{A}f(x) = \sum_{i=1}^n \mu_t^i \frac{\partial f}{\partial x^i}
    + \frac{1}{2}\operatorname{tr}\!\left[\sigma_t^\top H_f \sigma_t\right],
\end{equation}
where $H_f$ denotes the Hessian matrix with entries 
$[H_f]_{ij} = \frac{\partial^2 f}{\partial x^i \partial x^j}$, 
Itô's formula takes the compact form
\begin{equation}
    df = \left(\partial_t f + \mathcal{A}f\right)dt 
    + \sum_{i=1}^n \frac{\partial f}{\partial x^i}\sigma_t^i\, dW_t.
\end{equation}
In particular, $f(t, X_t)$ is a martingale if and only if 
$\partial_t f + \mathcal{A}f = 0$, which is a partial differential 
equation in $(t,x)$ whose coefficients are exactly those of the 
SDE \eqref{eq:multiSDE}. This connection between the dynamics of $X$ 
and the theory of partial differential equations is the foundation 
of the subsequent analysis.
\end{remark}
}

\begin{remark}[Infinitesimal Generator]
\label{rem:generator}
The drift in \eqref{eq:Ito} decomposes naturally into an explicit 
time derivative $\partial_t f$ and a purely spatial part determined by 
$\mu_t$ and $\sigma_t$. Defining the latter as the \emph{infinitesimal generator}
\begin{equation}
    \mathcal{A}f(x) = \sum_{i=1}^n \mu_t^i \frac{\partial f}{\partial x^i}
    + \frac{1}{2}\operatorname{tr}\!\left[\sigma_t^\top H_f \sigma_t\right],
    \label{eq:generator}
\end{equation}
where $H_f$ denotes the Hessian matrix with entries 
$[H_f]_{ij} = \frac{\partial^2 f}{\partial x^i \partial x^j}$, 
Itô's formula takes the compact form
\begin{equation}
    df = \left(\partial_t f + \mathcal{A}f\right)dt 
    + \sum_{i=1}^n \frac{\partial f}{\partial x^i}\sigma_t^i\, dW_t.
\end{equation}
In particular, $f(t, X_t)$ is a martingale if and only if 
$\partial_t f + \mathcal{A}f = 0$, a partial differential 
equation in $(t,x)$ whose coefficients are exactly those of the 
SDE \eqref{eq:multiSDE}. This connection between the dynamics of $X$ and the theory of partial differential equations is made precise by the Feynman--Kac formula of Section~\ref{subsec:fk}, and underlies the pricing framework developed in later chapters.
\end{remark}


\subsection{Existence and Uniqueness of Solutions}

We now establish conditions under which the SDE admits a unique solution. More precisely, we seek a process $X$ satisfying \eqref{eq:SDE} for all $t\geq0$.

\begin{proposition}
\label{prop:sde_existence}
    Suppose that there exists a constant K such that the following conditions are satisfied for all $x,y$ and $t$.
    \begin{align}
     ||\mu(t,x)-\mu(t,y)||&\leq K ||x - y||\\
     ||\sigma(t,x)-\sigma(t,y)||&\leq K ||x - y||\\
     ||\mu(t,x)||+||\sigma(t,x)||&\leq K (1+||x||)
    \end{align}
     Then the SDE \eqref{eq:SDE} admits a unique strong solution. The solution has the properties
    \begin{enumerate}
        \item $X$ is $\mathcal{F}_t^W$-adapted.
        \item $X$ has continuous trajectories.
        \item $X$ is a Markov process.
        \item \textcolor{red}{maybe not a use of the alst one?}
    \end{enumerate}
\end{proposition}

The first two conditions are \textit{Lipschitz conditions} on $\mu$ and $\sigma$ respectively, ensuring the coefficients do not vary too \textit{wildly} in space. The third is a \textit{linear growth condition} preventing the solution from blowing up in finite time. Notably, the solution inherits the Markov property. 

\begin{definition}[Markov Property(\textcolor{red}{insert reference})]
\label{def:markov}
    A stochastic process $X_t$ is said to be a Markov process if for all 
    $s \leq t$ and all measurable sets $A$:
    \begin{equation}
        P(X_t \in A \mid \mathcal{F}_s) = P(X_t \in A \mid X_s).
    \end{equation}
\end{definition}

This property is fundamental to the pricing framework developed later, as it ensures that expectations of functions of $X$ depend only on the current state $X_s$ rather than on the full history of the process.

\subsection{The Feynman--Kac Formula}
\label{subsec:fk}

Remark~\ref{rem:generator} established one direction of the link between diffusion processes and partial differential equations: if $f(t,X_t)$ is a martingale, then its drift must vanish, which leads to the equation
\[
\partial_t f + \mathcal{A}f = 0.
\]
In financial applications, the converse direction is particularly important. Prices are typically expressed as conditional expectations under a pricing measure, and one then seeks the differential equation satisfied by that expectation. For Markov diffusion processes, the Feynman--Kac formula provides precisely this connection between conditional expectations and parabolic PDEs.

We state a version that allows for a zeroth-order term in the PDE. This is the form relevant for asset-pricing applications, where the quantity of interest is a discounted conditional expectation and the discounting enters through the additional term $-k\,u$.

\begin{theorem}[Feynman--Kac, {\citealp[Prop.~5.6]{Bjork2020}}]
\label{thm:fk}
Let $\varphi : \mathbb{R}^n \to \mathbb{R}$ be continuous and satisfy a polynomial growth condition, let $k : [0,T] \times \mathbb{R}^n \to \mathbb{R}$ be bounded and continuous, and suppose that
\[
u \in C([0,T]\times\mathbb{R}^n)\cap C^{1,2}([0,T)\times\mathbb{R}^n)
\]
solves the terminal-value problem
\begin{equation}
\begin{cases}
    \partial_t u(t,x)+\mathcal{A}u(t,x)-k(t,x)\,u(t,x)=0,
    & (t,x)\in[0,T)\times\mathbb{R}^n, \\[4pt]
    u(T,x)=\varphi(x),
    & x\in\mathbb{R}^n,
\end{cases}
\label{eq:fk_pde}
\end{equation}
where $\mathcal{A}$ is the infinitesimal generator defined in \eqref{eq:generator}. Then $u$ admits the probabilistic representation
\begin{equation}
u(t,x)
=
\mathbb{E}\!\left[
    \exp\!\left(-\int_t^T k(s,X_s)\,ds\right)\varphi(X_T)
    \,\middle|\,
    X_t=x
\right],
\label{eq:fk_formula}
\end{equation}
where $X$ is the diffusion process with dynamics \eqref{eq:multiSDE}.
\end{theorem}

\begin{proof}
See \citet[Prop.~5.6]{Bjork2020}. A version under weaker regularity conditions is given in \citet[Thm.~5.3]{Friedman1975}.
\end{proof}

Two aspects of \eqref{eq:fk_formula} are particularly important. First, the representation depends on the past only through the current state $(t,x)$, reflecting the Markov property from Definition~\ref{def:markov}. Second, if $k\equiv 0$, then \eqref{eq:fk_formula} reduces to
\[
u(t,x)=\mathbb{E}\!\left[\varphi(X_T)\mid X_t=x\right],
\]
while \eqref{eq:fk_pde} reduces to
\[
\partial_t u+\mathcal{A}u=0,
\]
which is exactly the drift condition from Remark~\ref{rem:generator}. The Feynman--Kac formula can therefore be viewed as the natural extension of the martingale relation to the case where the conditional expectation is weighted by an exponential discount factor.

This result will be central in the pricing theory developed below. In particular, when the discount factor is generated by the short rate, the Feynman--Kac formula yields the PDE representation of arbitrage-free bond prices.

\section{Arbitragefree Pricing}

We consider a filtered probability space $(\Omega,\mathcal{F}, \{\mathcal{F}_t\}_{t\geq0},\mathbb{P})$, supporting a $d$-dimensional standard Brownian motion $W$ (\textcolor{red}{skal man her nævne noget med independen teller correlated?}). Our setup is as follows

We assume that there are $n$ risky assets $S^1,\dots,S^n$

under the objective probability measure $P$, the $S$-dynamics are given by the stochastic process we have in our stochasic calculus section. 

The coefficient processes are assumed to be adapted $\mu^i$ and $\sigma^{ij}$ above are assumed to be adapted, and we interpret them as the \textit{local mean rate of return} and \textit{volatility} of $S$ respectively. 

We have a standard risk free asset with price process $B$ with dynamics
$$dB_t=r_tB_tdt$$
where the short rate $r$ is assumed to be an adapted stochastic process. This will be interpreed as the bank accoutn with the interpreation that ..., and this is chosen to be our numeraire.
\textcolor{red}{we must be careful with the domestic measure here.}



Note:
we do not assume that we have the same number of Brownian motions driving the number of risky assets. and the probability space is also allowed to carry out other processes than the Brownian motion, such that we allow $\mathcal{F}_t^W\subseteq\mathcal{F}_t$ implying that the filtration could be generated by other processes beside $W$. For example, we make no assumptions about the distribution of the short rate process $r$ or of the processes $\mu_i$ and $\sigma_{ij}$, we only assume that they are adapted. In particular, this allows for models where these processes are path dependent upon the Brownian motion, or that they are driven by som other "hidden" state variable processes.







We consider a financial market consisting of a risk-free asset defined as the bank account and will operate as our numerair and $N$ risky assets.

and then we want to in this setup define state the first fundamental theorem when this market model is arbitrage free. 



pricing contingent claims. where the task is to  determining a "reasonable" price process assuming the primary market is arbitrage free.

note that different choices of $Q$ will generically give rise to different price processes. in 11.41, however assuming that the numeraire is the bank account then this gives rise to the 
risk neutral valuation formula.



Let the money-market account be defined by
\begin{equation}
B(t)
=
\exp\!\left(
\int_0^t r(u)\,du
\right),
\label{eq:bank_account}
\end{equation}
where $r(t)$ denotes the instantaneous short rate.

Under the risk-neutral measure $\mathbb{Q}$, the discounted price process of any traded asset is a martingale. In particular, the arbitrage-free price of a default-free zero-coupon bond satisfies
\begin{equation}
P(t,T)
=
\mathbb{E}_t^{\mathbb{Q}}
\left[
\frac{B(t)}{B(T)}
\right]
=
\mathbb{E}_t^{\mathbb{Q}}
\left[
\exp\!\left(
-\int_t^T r(u)\,du
\right)
\right].
\label{eq:riskneutral_bond}
\end{equation}



Once the short-rate dynamics under $\mathbb{Q}$ are specified, bond prices, and thereby the entire discount curve, are determined.

To ready us selved for interest rate theory (in björk chap. 19-24) we need to hereby state som elementary concests

Def. 11.2 a ZCB

and its price 11.21

\newpage

\section{Arbitrage-Free Pricing}

We now turn to the financial setting in which the stochastic-calculus results above are used. The fundamental idea is that, in an arbitrage-free market, suitably discounted asset prices are martingales under an equivalent martingale measure. As a consequence, contingent claims can be valued by conditional expectation. When the relevant state variables form a Markov diffusion, the Feynman--Kac formula provides the associated pricing PDE.

Consider a filtered probability space $(\Omega,\mathcal{F},\{\mathcal{F}_t\}_{t\geq 0},\mathbb{Q})$ supporting a $d$-dimensional Brownian motion $W$. We assume that the market consists of a money-market account and a collection of traded assets. The money-market account is defined by
\begin{equation}
B_t=\exp\!\left(\int_0^t r_s\,ds\right),
\label{eq:bank_account}
\end{equation}
where $r_t$ denotes the short-rate process. We take $B_t$ as numeraire and work under an equivalent martingale measure $\mathbb{Q}$ associated with this choice of numeraire.

Under $\mathbb{Q}$, the discounted price process of any traded asset is a martingale. More generally, if a contingent claim pays $H$ at time $T$, then its arbitrage-free price process is given by the risk-neutral valuation formula
\begin{equation}
\Pi_t
=
\mathbb{E}^{\mathbb{Q}}\!\left[
\exp\!\left(-\int_t^T r_s\,ds\right)H
\,\middle|\,
\mathcal{F}_t
\right].
\label{eq:riskneutral_valuation}
\end{equation}

A particularly important special case is the default-free zero-coupon bond, which pays one unit of currency at maturity $T$. Its arbitrage-free price is therefore
\begin{equation}
P(t,T)
=
\mathbb{E}^{\mathbb{Q}}\!\left[
\exp\!\left(-\int_t^T r_s\,ds\right)
\,\middle|\,
\mathcal{F}_t
\right].
\label{eq:riskneutral_bond}
\end{equation}

Equation \eqref{eq:riskneutral_bond} shows that once the short-rate dynamics under $\mathbb{Q}$ are specified, zero-coupon bond prices are determined. Since bond prices form the basic building blocks of fixed-income markets, we now introduce the main interest-rate quantities derived from them.


\newpage

\section{Interest Rate Theory}

The fundamental objects in fixed-income markets are zero-coupon bonds, since the entire discount curve is generated by the family of bond prices $\{P(t,T)\}_{T\geq t}$. From these prices one may derive yields, forward rates, and swap rates.

\begin{definition}[\textit{Zero Coupon Bond}]
A zero coupon bond with maturity date $T$, also called a $T$-bond, is a contract which guarantees the holder one unit of currenc at time $T$. The price at time $t\leq T$ of a $T$-bond is denoted by $P(t,T)$, with the terminal condition
$$P(T,T)=1.$$
\end{definition}

We assume the existence of a \textcolor{red}{frictionless} and arbitrage-free bond market for all maturities $T>0$, and that pricing is performed under a risk neutral martingale measure $Q$. Under $Q$, the price of a $T$-bond is given by the risk neutral valuation formula

\begin{equation}
    P(t,T)=E^Q\left[e^{-\int_t^T r_s\,ds}\,\middle|\,\mathcal{F}_t\right]
    \label{ZCBprice}
\end{equation}
where $r_t$ denotes the short rate process governing the instantaneous cost of borrowing at time $t$. 

\begin{assumption}[\textit{Single Curve}]
A single discount curve, determined entirely by the ZCB prices $\{P(t,T)\}_{T\geq t}$, is used for both discountig and forwarding. Under this assumption, the forward rate over the accrual period $[T_{i-1},T_i]$ satisfies
$$1+\Delta_i L(t;T_{i-1},T_i)=\frac{P(t,T_{i-1})}{P(t,T_i)},$$
where $\Delta_i=T_i-T_{i-1}$ denotes the accrucal fraction.
\label{as:singlecurve}
\end{assumption}


Pr. Assumption \ref{as:singlecurve}, we can extract the yield curve from the ZCB prices, which describes the relationship between interest rates and maturities at a given point in time.

\begin{definition}[\textit{Yield}]
    The continuously compounded yield $R(t,T)$ of a $T$-bond at time $t$ is defined by 
    \begin{equation}
        P(t,T)=e^{-R(t,T)(T-t)} 
        \label{eq:yield1}
    \end{equation}
    or equivalently
    \begin{equation}
        R(t,T)=-\frac{\log P(t,T)}{T-t} 
        \label{eq:yield2}
    \end{equation}
\end{definition}
The single curve assumption ensures that one can move freely between the ZCB price $P(t,T)$ and yield $R(t,T)$ through \eqref{eq:yield1}, and that entire term structure is determined by a single discount curve. \textcolor{red}{The yield curve is the primary object we aim to model throughout this thesis.}

\textcolor{red}{what about introducing forward rates as well?}

\begin{definition}[\textit{Instantaneous Forward Rate}]
The instantaneous forward rate for maturity $T$ observed at time $t$ is defined by
\begin{equation}
f(t,T)=-\frac{\partial}{\partial T}\log P(t,T),
\label{eq:inst_forward}
\end{equation}
provided the derivative exists.
\end{definition}

The forward-rate curve provides an equivalent representation of the term structure. In particular, bond prices can be recovered from forward rates through
\begin{equation}
P(t,T)=\exp\!\left(-\int_t^T f(t,u)\,du\right).
\label{eq:bond_from_forward}
\end{equation}

For a discrete accrual period $[T_{i-1},T_i]$, the corresponding simply compounded forward rate is defined by
\begin{equation}
1+\Delta_i L(t;T_{i-1},T_i)=\frac{P(t,T_{i-1})}{P(t,T_i)},
\label{eq:discrete_forward}
\end{equation}
where $\Delta_i=T_i-T_{i-1}$ denotes the accrual fraction.


\subsection{Interest Rate Swaps}

With the interest rate objects in place. We can now continout to introduce the simplest of all interest rate derivivatives namely the interest rate swap and also the derivation of its price.

This is the scheme of exchanging a payment stream at a fixed interest, known as the \textbf{swap rate}, for a payment stream at a floating rate. Being many version os interest rate swaps we assess ...


Consider a fixed-for-floating interest rate swap initiated at time $t$ and maturing at $T=t+\tau$. Let
\[
t<T_1<\dots<T_n=T
\]
denote the payment dates and define the accrual fractions
\[
\Delta_i = T_i-T_{i-1}, \qquad i=1,\dots,n.
\]

Let $N$ denote the notional. If the fixed rate of the swap is $S(t,\tau)$, the value of the fixed leg at time $t$ is
\begin{equation}
V_{\mathrm{fix}}(t)
=
N S(t,\tau)
\sum_{i=1}^{n}\Delta_i P(t,T_i).
\label{eq:fixed_leg}
\end{equation}

Under the single-curve assumption adopted throughout this thesis, the forward rate over the accrual period $[T_{i-1},T_i]$ satisfies
\[
1+\Delta_i L(t;T_{i-1},T_i)
=
\frac{P(t,T_{i-1})}{P(t,T_i)}.
\]

For a standard spot-starting swap, the value of the floating leg follows from the usual telescoping replication argument and is given by
\begin{equation}
V_{\mathrm{float}}(t)
=
N\bigl(1-P(t,T)\bigr).
\label{eq:floating_leg}
\end{equation}

The par swap rate is defined as the fixed rate that sets the value of the swap equal to zero. Solving
\[
V_{\mathrm{float}}(t)-V_{\mathrm{fix}}(t)=0
\]
yields
\begin{equation}
S(t,\tau)
=
\frac{1-P(t,T)}
{\sum_{i=1}^{n} \Delta_i P(t,T_i)}.
\label{eq:swap_rate}
\end{equation}

Hence, par swap rates are deterministic nonlinear functions of the underlying discount curve. Once an arbitrage-free discount curve is specified, the corresponding swap curve is determined through \eqref{eq:swap_rate}.



\subsection{To be incporated}


NOTE:

Chapter 14 in björk?


Forward Rate Models (chapter 5, Björk)

Notes to incorporate:
It is practical to use time \textit{to} maturity rather than time \textit{of} maturity, to parameterize the forward rates. The point of this parametrization is that it highlights the forward rate dynamics as an infinite dimensional SDE. Making it an indespensable tool of modern interest rate theory.


The pricing relations above are model-free consequences of arbitrage-free fixed-income theory. To obtain a tractable dynamic model of the yield curve, one must specify the evolution of the underlying state variables under the risk-neutral measure. A particularly important and analytically convenient class is provided by affine term structure models.

%------------------------------------------
\subsection{Affine Term Structure Models}
\label{sec:Affine TSM Theory}
%------------------------------------------

In general term structure models will be introduced here (Chap. 22 (p. 298)) where we specifically come with AFfine as example. Markovian factor models


Affine term structure models provide a tractable way of representing the term structure in an arbitrage-free setting. Their central idea is that bond prices are driven by a finite-dimensional factor process rather than by an unrestricted infinite-dimensional curve.

Let $X_t \in \mathbb{R}^m$ denote an $m$-dimensional factor process under the risk-neutral measure $\mathbb{Q}$. In an affine term structure model, the short rate is assumed to be affine in the factors,
\begin{equation}
r(t) = \delta_0 + \delta_1^\top X_t,
\label{eq:atsm_short_rate}
\end{equation}
where $\delta_0 \in \mathbb{R}$ and $\delta_1 \in \mathbb{R}^m$.

The defining property of an affine term structure model is that zero-coupon bond prices take an exponential-affine form in the state variables:
\begin{equation}
P(t,T)
=
\exp \bigl(
A(T-t) + B(T-t)^\top X_t
\bigr),
\label{eq:affine_form}
\end{equation}
where $A(\tau) \in \mathbb{R}$ and $B(\tau) \in \mathbb{R}^m$ are deterministic functions of time to maturity $\tau=T-t$.

Equivalently, in Musiela parametrization,
\begin{equation}
P(t,t+\tau)
=
\exp \bigl(
A(\tau) + B(\tau)^\top X_t
\bigr).
\label{eq:affine_form_musiela}
\end{equation}

This representation implies that the term structure is generated by the factor vector $X_t$. Affine models are therefore attractive both theoretically and empirically, since they combine a structured state representation with arbitrage-free bond pricing.

Under suitable affine specifications of the factor dynamics under $\mathbb{Q}$, the exponential-affine bond-pricing form leads to a system of ordinary differential equations for the coefficient functions $A(\tau)$ and $B(\tau)$. 


