%=========================================================
\chapter{Simulation-Based Swaption Pricing and Calibration}
\label{chap:pricing_methodology}
%=========================================================

The previous chapter established that the encoder--decoder model produces stable latent dynamics suitable for scenario generation. We now ask a harder question: can the same model be used to price swaptions? This requires the \emph{risk-neutral} distribution of future swap rates, which is a fundamentally different object from the historical cross-section the model was trained on.

%=========================================================
\section{Monte Carlo Swaption Pricing}
\label{sec:swaption_pricing_methodology}
%=========================================================

We consider European swaptions \citep{Filipovic2009} written on the single-curve swap rates introduced in Section~\ref{subsec:interest_rate_swaps}. The option expires at \(T_e\) and gives the holder the right to enter a swap running from \(T_e\) to \(T_e+n\), with annual fixed-leg payment dates \(T_e+1, T_e+2, \ldots, T_e+n\).

For an at-the-money (ATM) swaption the strike equals the time-zero forward swap rate. Given the discount curve \(P(0,T)\), the forward annuity and forward swap rate are
\[
\mathcal A_0
=
\sum_{j=1}^{n}P(0,T_e+j),
\qquad
F_0
=
\frac{P(0,T_e)-P(0,T_e+n)}{\mathcal A_0},
\]
so the ATM strike is \(K = F_0\).

On each Monte Carlo path \(m=1,\ldots,N_{\mathrm{MC}}\), the latent state is simulated to the option expiry \(T_e\) using a fine time grid with step \(dt \ll 1\). The decoder then produces the time-\(T_e\) zero-coupon bond curve \(\tau\mapsto P^{(m)}(T_e,\tau)\), from which the pathwise annuity and swap rate are
\[
\mathcal A^{(m)}(T_e)
=
\sum_{j=1}^{n}P^{(m)}(T_e,j),
\qquad
S^{(m)}(T_e)
=
\frac{1-P^{(m)}(T_e,n)}
{\mathcal A^{(m)}(T_e)}.
\]
The payer and receiver swaption payoffs are
\[
X^{(m),\mathrm{pay}}_{T_e}
=
\mathcal A^{(m)}(T_e)\bigl(S^{(m)}(T_e)-K\bigr)^+,
\qquad
X^{(m),\mathrm{rec}}_{T_e}
=
\mathcal A^{(m)}(T_e)\bigl(K-S^{(m)}(T_e)\bigr)^+,
\]
where we set notional \(\mathcal N = 1\). Each payoff is discounted by the pathwise money-market factor
\[
D^{(m)}_{T_e}
=
\exp\!\left(-\int_0^{T_e}r_s^{(m)}\,ds\right),
\]
approximated on the simulation grid by the trapezoidal rule (Chapter~\ref{chap:simulation}). The Monte Carlo price estimator \citep{Glasserman2004} is
\[
\widehat V_0
=
\frac{1}{N_{\mathrm{MC}}}
\sum_{m=1}^{N_{\mathrm{MC}}}
D^{(m)}_{T_e}X^{(m)}_{T_e}.
\]

\subsection{Bachelier implied normal volatility}
\label{subsec:bachelier_formula}

Market swaption quotes are reported as normal (Bachelier \citep{Bachelier1900}) implied volatilities \(\sigma_N\). Under the Bachelier model the forward swap rate satisfies \(dF(t)=\sigma_N\,dW^{\mathcal A}(t)\) under the annuity measure \citep[Ch.~26]{Bjork2020}, giving the ATM price
\[
V^{\mathrm{ATM}}
=
\mathcal A_0
\frac{\sigma_N\sqrt{T_e}}{\sqrt{2\pi}}.
\]
The normal convention is standard in low- and negative-rate environments because it permits negative rates \citep{Antonov2015}. To convert a Monte Carlo price into an implied vol we invert this relation.

To decouple the vol level from any directional bias, we price both a payer and a receiver at the ATM strike and average:
\[
\widehat V_{\mathrm{str}}
=
\tfrac{1}{2}\!\left(
\widehat V^{\mathrm{pay}}_{\mathrm{MC}}
+
\widehat V^{\mathrm{rec}}_{\mathrm{MC}}
\right),
\qquad
\widehat{\sigma}_{N}
=
\frac{\widehat V_{\mathrm{str}}\sqrt{2\pi}}
{\mathcal A_0\sqrt{T_e}}.
\]
This \emph{straddle vol} eliminates the effect of the distribution being off-centre: if the model shifts the forward rate up or down, both payer and receiver prices move in opposite directions and the average is unchanged at first order. All vols are reported in basis points (\(\times 10^4\)).

We also track the \emph{forward centering bias}
\[
\widehat b
=
\frac{
\widehat V^{\mathrm{pay}}_{\mathrm{MC}}
-
\widehat V^{\mathrm{rec}}_{\mathrm{MC}}
}
{\mathcal A_0},
\]
which should be close to zero at the ATM strike for a correctly centred model.

%=========================================================
\section{Pricing with the Stable Model}
\label{sec:pricing_stable_model}
%=========================================================

The stable model from Chapter~\ref{chap:model} is written entirely under the risk-neutral measure \(\mathbb{Q}\),
\[
dz_t = K(z_t)\,dt + L(z_t)\,dW_t^{\mathbb{Q}},
\]
and the no-arbitrage condition is enforced by construction through the bond-pricing ODE. In principle this is all that is needed to price options: simulate latent paths under the stable dynamics, decode to bond curves at expiry, and compute the Monte Carlo estimator of Section~\ref{sec:swaption_pricing_methodology}. There is no measure change to perform.

\subsection{Why this is not enough}
\label{subsec:stable_pricing_issue}

The difficulty is a calibration mismatch. The parameters of \(K\), \(H\), and \(G\) were chosen to minimise the curve reconstruction loss
\[
\mathcal{L}_{\mathrm{recon}}
= \mathbb{E}\!\left[\bigl\|S - \hat S(z)\bigr\|^2\right],
\]
which measures how accurately the decoder reproduces today's observed swap curve. This objective contains no information about the \emph{distribution of future} swap rates and no information about option markets. A drift and diffusion that fit curve cross-sections well are not guaranteed to reproduce the option-implied risk-neutral distribution of \(F_{T_e}\).

Concretely, two problems can arise:

\begin{enumerate}
    \item \textbf{Wrong volatility level.} The diffusion \(L(z)\) and drift \(K(z)\) together determine the spread of the distribution of \(F_{T_e}\) under \(\mathbb{Q}\). Since neither was calibrated against option prices, the resulting implied vol can differ substantially from market-observed levels.
    \item \textbf{Wrong forward centering.} The drift \(K(z)\) is calibrated to reconstruct the current curve, not to centre the simulated forward swap rate at \(F_0\). Paths drift away from the ATM forward, producing a large forward bias \(\widehat b \neq 0\).
\end{enumerate}

Whether these are mild or severe is an empirical question.

\subsection{Empirical results}
\label{subsec:stable_pricing_results}

We price all available EUR ATM swaption cells using the stable model directly, without any adjustment. The simulation uses \(N_{\mathrm{MC}} = 512\) antithetic paths per cell and the time grid described in Section~\ref{sec:swaption_pricing_methodology}.

\begin{table}[H]
\centering
\input{Figures/Pricing/eval_base/tab_base_per_cell_all.tex}
\end{table}

The overall MAE is \textbf{414.7 bp}, far exceeding typical bid--offer spreads of 1--5 bp. Short-expiry cells are the worst: the 1Yx5Y cell exceeds 1000 bp. All cells show a large positive vol error — the stable model systematically overestimates implied volatility — and a strongly negative forward bias, confirming that the simulated forward distributions are not centred at \(F_0\).

This is the central diagnostic of this chapter: a latent-state model that reconstructs curves accurately is not, by itself, a swaption-pricing model. The two problems anticipated above are both present in extreme form. The remainder of the chapter introduces a small trainable pricing layer on top of the frozen stable model, calibrated against observed swaption vols, that closes most of this gap.

%=========================================================
% --- Subsequent sections (pricing layer + comparison) will be added once the
%     final model choice has been made.  Candidates currently under evaluation:
%
%       Constant MPR
%       Expiry MPR
%       Expiry-Tenor Vol MPR
%       State-Conditioned Vol MPR (saddle-point fix re-run)
%       SV Pricing MPR (CIR stochastic vol with state-conditioned v_0)
%       Surface Vol MPR (full LOO cross-section)
%
%     Once the chosen model and its results are confirmed, the next section
%     will introduce the architecture, calibration objective, and present
%     per-cell results and comparison against the stable-model baseline above.
%=========================================================
