\subsection{Stabilization of the drift network $K$}

In the baseline specification, the drift network $K$ is modeled as a linear map
\begin{equation}
\mu(z)=Mz+N,
\label{eq:k_baseline}
\end{equation}
where $M \in \mathbb{R}^{d \times d}$ and $N \in \mathbb{R}^d$ are trainable parameters.
This formulation is simple and expressive, but it does not guarantee stable latent
dynamics under forward simulation. In particular, if the learned matrix $M$ has one or
more eigenvalues with non-negative real part, the deterministic part of the state
dynamics may fail to be mean-reverting, which can lead to unstable or explosive sample
paths.

To address this issue, we replace the unconstrained matrix $M$ by the structured
parameterization
\begin{equation}
M = -\left(V^\top V + \varepsilon I_d\right),
\label{eq:k_stable_matrix}
\end{equation}
where $V \in \mathbb{R}^{d \times d}$ is a trainable matrix, $I_d$ is the identity matrix
in dimension $d$, and $\varepsilon > 0$ is a small constant. The drift network is then
given by
\begin{equation}
\mu(z)= -\left(V^\top V + \varepsilon I_d\right)z + N.
\label{eq:k_stable_drift}
\end{equation}

This reparameterization guarantees that the drift matrix is negative definite. Since
$V^\top V$ is positive semidefinite, the matrix $V^\top V + \varepsilon I_d$ is strictly
positive definite for every $\varepsilon > 0$. Hence, for any nonzero vector
$x \in \mathbb{R}^d$,
\begin{equation}
x^\top M x
=
- x^\top \left(V^\top V + \varepsilon I_d\right)x
=
-\|Vx\|^2 - \varepsilon \|x\|^2
< 0.
\end{equation}
It follows that all eigenvalues of $M$ are real and strictly negative. The linear drift
therefore has an inherent pull toward an equilibrium level and cannot generate locally
explosive deterministic behavior.

The equilibrium point is obtained by solving $\mu(z^\ast)=0$, which yields
\begin{equation}
z^\ast = \left(V^\top V + \varepsilon I_d\right)^{-1}N.
\end{equation}
Equivalently, the drift can be written as
\begin{equation}
\mu(z)=-Q(z-z^\ast),
\qquad
Q:=V^\top V+\varepsilon I_d,
\end{equation}
which makes the mean-reverting structure explicit: deviations from the equilibrium state
$z^\ast$ are pulled back at a rate governed by the positive definite matrix $Q$.

The motivation for this modification was numerical instability observed when simulating
from the original unconstrained specification in \eqref{eq:k_baseline}. In that case,
stability depends entirely on the outcome of the optimization, since the learned matrix
$M$ may contain positive or weakly damped eigenvalues. By contrast, the parameterization
in \eqref{eq:k_stable_matrix} guarantees a stable linear drift for every value of the
trainable matrix $V$. This is particularly desirable in the present setting, where the
latent factors enter a continuous-time term-structure model and must remain well behaved
under forward simulation.

An additional advantage is that the stabilized parameterization does not increase the
number of trainable parameters. The original formulation uses $d^2$ parameters for $M$
and $d$ parameters for $N$, while the stabilized version uses $d^2$ parameters for $V$
and $d$ parameters for $N$. Thus, both specifications contain $d^2+d$ trainable
parameters. The extra matrix multiplication required to form $V^\top V$ is of the same
asymptotic order as the original linear mapping and is negligible in the present low-
dimensional setting.

In the implementation, $V$ is initialized orthogonally and $N$ is initialized at zero.
Orthogonal initialization provides a well-scaled starting point for optimization, while
the constant $\varepsilon$ ensures strict negative definiteness even in directions where
$V^\top V$ is close to singular. The resulting network preserves the linear form of the
original drift specification, but enforces stable and mean-reverting latent dynamics by
construction.

\subsection{Stabilization of the short-rate network $R$}

In the baseline specification, the short-rate network $R$ maps the latent factors $z$ to
the instantaneous short rate through a feedforward neural network,
\begin{equation}
\tilde r(z) = R_\theta(z),
\label{eq:r_baseline}
\end{equation}
where $R_\theta : \mathbb{R}^d \to \mathbb{R}$ is implemented as a two-layer network with
centered soft step activation. This specification is flexible, but it leaves the network
output unconstrained. As a result, the learned short rate may attain implausibly large
positive or negative values outside the region well covered by the training data. Such
behavior is undesirable in a continuous-time term-structure model, where the short rate
enters both the discount factor and the coefficients of the no-arbitrage ODE system.

To improve numerical robustness, we replace the unconstrained output layer by a bounded
parameterization. Specifically, we write
\begin{equation}
\tilde r(z) = r_{\mathrm{center}} + r_{\mathrm{scale}} \tanh\!\bigl(f_\theta(z)\bigr),
\label{eq:r_stable}
\end{equation}
where $f_\theta : \mathbb{R}^d \to \mathbb{R}$ is a neural network with the same hidden
architecture as the original specification, $r_{\mathrm{center}} \in \mathbb{R}$ is a
learnable location parameter, and $r_{\mathrm{scale}} > 0$ is a learnable scale parameter.
To ensure positivity of the scale, we parameterize it as
\begin{equation}
r_{\mathrm{scale}} = \mathrm{softplus}(a) + \varepsilon_r,
\label{eq:r_scale}
\end{equation}
where $a \in \mathbb{R}$ is unconstrained and $\varepsilon_r > 0$ is a small constant.
This guarantees that the scale remains strictly positive throughout training.

The representation in \eqref{eq:r_stable} implies the bound
\begin{equation}
\tilde r(z) \in
\left(
r_{\mathrm{center}} - r_{\mathrm{scale}},
\;
r_{\mathrm{center}} + r_{\mathrm{scale}}
\right),
\end{equation}
since $\tanh(x) \in (-1,1)$ for all $x \in \mathbb{R}$. Hence, the short rate is bounded
by construction. Unlike hard clipping, this transformation is smooth and differentiable,
which makes it more suitable for gradient-based optimization. In particular, it preserves
continuous gradients while still preventing the network from producing extreme outputs.

The motivation for this modification was numerical instability observed when simulating
from the unconstrained short-rate network. Because $\tilde r(z)$ enters the discount
factor through
\begin{equation}
B(t) = \exp\!\left(\int_0^t \tilde r(z_u)\,du\right),
\end{equation}
large values of the short rate can lead to highly unstable discounting behavior and may
also amplify errors in the ODE coefficients used in the decoder. By contrast, the bounded
specification in \eqref{eq:r_stable} keeps the short rate within a learnable but controlled
range, thereby improving the robustness of both training and forward simulation.

An important feature of this construction is that it does not impose a fixed hard-coded
interval for the short rate. Instead, both the center and the scale are learned from the
data. The network can therefore adapt the admissible range to the prevailing rate
environment. In particular, the model is still able to represent negative short rates when
supported by the data, which is important in modern interest-rate markets. The purpose of
the transformation is thus not to enforce non-negativity, but to avoid implausibly large
magnitudes and to keep the model numerically well behaved.

In the implementation, the hidden network retains the same structure as the original
short-rate network, while the final output is transformed through the bounded mapping in
\eqref{eq:r_stable}. This preserves the flexibility of the original architecture in the
interior of the state space, but adds a structural safeguard at the output level. The
resulting specification is therefore better suited for stable simulation and estimation in
the arbitrage-free term-structure framework.